---
title: "EDS241: Assignment 3"
author: "Wylie Hampson"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)


# load packages
packages=c("stargazer", "here", "tidyr", "dplyr","stringr", "janitor", "here",
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble", "estimatr")

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

#devtools::install_github('rstudio/rmarkdown')
options(scipen=999) # not scientific notation


```

\noindent
**Question 1:** Application of estimators based on treatment ignorability This exercise asks you to implement some of the techniques presented in Lectures 6-7. The goal is to estimate the causal effect of maternal smoking during pregnancy on infant birth weight using the treatment ignorability assumptions. The data are taken from the National Natality Detail Files, and the extract “SMOKING_EDS241.csv”' is a random sample of all births in Pennsylvania during 1989-1991. Each observation is a mother-infant pair. The key variables are:

\noindent
**The outcome and treatment variables are:**
birthwgt=birth weight of infant in grams tobacco=indicator for maternal smoking 

\noindent
**The control variables are:** 
mage (mother's age), meduc (mother's education), mblack (=1 if mother black), alcohol (=1 if consumed alcohol during pregnancy), first (=1 if first child), diabete (=1 if mother diabetic), anemia (=1 if mother anemic)

\noindent
**Import the data:**

```{r}
smoking <- read.csv(here("data", "SMOKING_EDS241.csv"))
```

\noindent
**Question a:** What is the unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers? Under what hypothesis does this correspond to the average treatment effect of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical evidence for or against this hypothesis.

```{r}
smoking_group <- smoking %>% 
  group_by(tobacco) %>% 
  summarize(mean_birthwgt = mean(birthwgt))

smoking_group

non_smoke_mean_wgt <- smoking_group %>% filter(tobacco == 0)
smoke_mean_wgt <- smoking_group %>% filter(tobacco == 1)

mean_wgt_diff <- non_smoke_mean_wgt$mean_birthwgt - smoke_mean_wgt$mean_birthwgt
```

\noindent
*The unadjusted mean difference in birth weight between babies from smoking and nonsmoking mothers is `r mean_wgt_diff` grams. This leads us to the hypothesis that on average maternal smoking causes a decrease in birth weights. The simple empirical evidence is the simple fact that on average babies from mothers who smoked had lower birth weights than those that did not smoke.*

\noindent
**Question b:** Assume that maternal smoking is randomly assigned conditional on the observable covariates listed above. Estimate the effect of maternal smoking on birth weight using a linear regression. Report the estimated coefficient on tobacco and its standard error.

```{r}
model_1 <- lm_robust(formula = birthwgt ~ tobacco, data = smoking)

model_1_table <- broom::tidy(model_1) %>% 
  dplyr::select(term, estimate, std.error, p.value) %>%
  knitr::kable()

model_1_table
```

\noindent
**Question c:** Use the exact matching estimator to estimate the effect of maternal smoking on birth weight. For simplicity, consider the following covariates in your matching estimator: create a 0-1 indicator for mother's age (=1 if mage>=34), and a 0-1 indicator for mother's education (1 if meduc>=16), mother's race (mblack), and alcohol consumption indicator (alcohol). These 4 covariates will create 2x2x2x2 = 16 cells. Report the estimated average treatment effect of smoking on birthweight using the exact matching estimator and its linear regression analogue (Lecture 6, slides 12-14).

```{r}
smoking <- smoking %>% 
  mutate(age_ind = ifelse((mage >= 34), 1, 0),
         edu_ind = ifelse((meduc >= 16), 1, 0))
```

\noindent
**Question d:** Estimate the propensity score for maternal smoking using a logit estimator and based on the
following specification: mother’s age, mother’s age squared, mother’s education, and indicators
for mother’s race, and alcohol consumption. 

```{r}
ps_model <- glm(tobacco ~ mage + (mage ^ 2) + meduc + mblack + alcohol, family = binomial(), data = smoking)
summary(ps_model)
EPS <- predict(ps_model, type = "response")
PS_WGT <- (smoking$tobacco/EPS) + ((1-smoking$tobacco)/(1-EPS))
```


\noindent
**Question e:** Use the propensity score weighted regression (WLS) to estimate the effect of maternal
smoking on birth weight (Lecture 7, slide 12).

```{r}
wls1 <- lm(formula = birthwgt ~ tobacco, data = smoking, weights = PS_WGT)

se_models = starprep(wls1, stat = c("std.error"), se_type = "HC2", alpha = 0.05)
stargazer(wls1, se = se_models, type="text")
```
















